Smoothing seismicity techniques applied to seismic source characterization and probabilistic seismic hazard analysis in Brazil

### 2015-05-14 manuscript.tex
data section almost reviewed. waiting mfd new figure

Brazil occupies a large stable continental region with seismicity rate lower than Central and Eastern North America, Australia, India or Northern Europe \citep{giardini_1999,johnston_1996a}. Spanning most of the central region of the South American plate, Brazil has earthquakes above magnitude 6 mb with a return period of 50 years or so. Historical records are very short with almost no data before the XX century. Historical evidence of magnitude a 7 has been reported in the Amazon \citep{berrocal_1984,veloso_2014}. However, as in other intraplate settings, seismic hazard is not negligible due to the low attenuation in the cold and thick lithosphere (magnitude 5 mb can be felt up to ~500 km away \citep{assumpcao_et_al_2014}) and special seismic hazard studies have been required for critical facilities such as nuclear power plants \citep{almeida_etal_2013}.

Few studies related to seismic hazard, such as \citet{berrocal_1996,santos_2010} for SE and NE Brazil, have been carried out so far and the first Brazilian seismic building code \citep{nbr_15421_2006,santos_lima_2008} was based on the GSHAP \citep{giardini_1999} results for South America \citep{tanner_shepherd_1997, shedlock_tanner_1999}. Brazilian seismicity is concentrated in several seismic zones \citep{berrocal_1984, assumpcao_et_al_2014}. Except for a peak of activity in Northeast Brazil, both the GSHAP model and the \citet{nbr_15421_2006} zoning (even when the Brazilian recommended levels are in line with those of other nations with higher occurrence of earthquakes \citep{santos_2013}) have not taken into proper account some of the other highly active seismic zones, such as the Porto dos Gaúchos, Mato Grosso state, with recurrent activity since the occurrence of the largest recorded Brazilian event with magnitude 6.2 mb in 1955 \citep{barros_2009}, or the Central Amazon region where concentration of events in the last 50 years is consistent with the location of the possible historical earthquake of magnitude 7 \citep{veloso_2014}. For this reason, it is important to review and update the hazard estimates for Brazil, especially since the number of Brazilian earthquake records has been increased in the last three decades. %the increasing number of Brazilian earthquake records in the last three decades. 
%In addition, the results presented here are part of the \acrfull{gem} supported SARA project (South America Risk Assessment).

@@

Data on intraplate seismicity in Brazil is currently the Brazilian Seismic Bulletin maintained mainly by the Seismological Center of the University of São Paulo \citep{bsb_2014}. This catalog is a joint compilation from different sources.

@@

For events up to 1981, the catalog corresponds to \citet{berrocal_1984} which is a comprehensive compilation of historic and instrumental data with participation of Universidade de São Paulo (USP), Universidade de Brasilia (UnB), Universidade Federal do Rio Grande do Norte (UFRN), Observatório Nacional (ON) and Instituto de Pesquisas Tecnológicas do Estado de São Paulo (IPT). From 1982 to 1995 the information comes from the annual bulletins published by Revista Brasileira de Geofísica. Since 1995, the Bulletin is maintained mainly by USP under the same cooperation network in addition to Universidade Estadual Paulista (UNESP). Earlier versions of this catalog were used to compose the well-known \citet{ceresis_1985, ceresis_1995} catalogs.

@@

This catalog only cover crustal events. Andean subduction earthquakes deeper than 50 km in the Brazil-Peru border was discarded as well earthquakes from the Mid-Atlantic ridge. The epicenters of the catalog are shown in figure \ref{fig_seismicity}.

@@

Most magnitudes were computed as $mb$ type or the equivalent $mR$ \citep{assumpcao_1983} regional magnitude. For historical events magnitudes were estimated from macroseismic data (felt area and maximum intensity) \citep{berrocal_1984}. Since most \gls{gmpe} are based on $M_W$ moment scale magnitude, in this present study, we converted all $mb$ magnitudes to $M_W$ values following the guidelines on table \ref{tab:guidelines}. REFERENCIA PARA STEPHANE/ASSUMPCAO WORK IN PROGRESS. (Johnston)
% Other proxies for intraplate regions like \citet{scordilis_2006} was discarded by its magnitude range definition, over than 6, whilst almost magnitude values on \citet{bsb_2014} catalog have magnitude values lower than that.

@@

Figures \ref{fig_overview}b and \ref{fig_overview}c show an almost uniform distribution of day of the week, and a slightly trend to record earthquakes during the night. This could suggest some influence of daily noise level.

@@

In the case of Brazilian catalog the methods did not present large differences at the final number of earthquakes (fig. \ref{fig_decluster}a) and the Gardner-Knopoff/Uhrhammer combination was chosen to maximize the number of events.

The map of figure \ref{fig_decluster}b locates all clusters in the catalog. It is an \emph{a-priori} evidence of regions in Brazil where long earthquake series tend to occur more frequently.

@@

Brazil just recently was installed its permanent seismic network \citep{pirchiner_2011}. The number and quality of stations increased over time. The population density, previously concentrated along the coast increased in other regions changing the historical detectability.

There are many methodologies to assessment of the temporal and spatial completeness \citep{stepp_1972, mignan_woessner_2012, ogata_1993, wiemer_wyss_2000, cao_gao_2002, stucchi_2004, woessner_2005, mignan_2011, mignan_2012, vorobieva_2013, mignan_2013, nasir_2013, mignan_2014}, or explaining how to compute the real \gls{pde} from the waveforms \citep{schorlemmer_2008}, but the time and the effort to proceed this exhaustive study of the \glsdesc{sym:Mc}, despite its relevance, was not intended to be done on the scope of this paper.

The waveforms from the earliest years was not in digital form. The histories of most regional (temporary) stations are not easy available. In addition, the low number of earthquakes in the catalog also difficults some high resolution statistical methods. All of this hinders the direct application of most that methodologies.

For this reason, the spatial completeness was not considered and the temporal completeness parameters for the whole country was defined by an easy cumulative rate criteria using historical information about the network evolution and the \citet{stepp_1972} plot (fig. \ref{fig_completeness}a). TALK ABOUT BRASILIA STATION???

Figure \ref{fig_completeness}b shows the magnitude time distribution in a scatter plot as the cumulative number of events which is important to perform the \gls{tcef} \citep{nasir_2013}. The linear behavior of cumulative rate on the recent years, gives an idea about the constant record trend. Figure \ref{fig_completeness}b also show three (from many) possible completeness tables to allow quick check their distinctions. We chosen the ``simplest'' one.

IMPROVE THIS DECISION EXPLANATION

%------------------------------------------------
%
% smooothing methods
%
%------------------------------------------------
\section{Smoothing methods}
Three smoothing methods are now briefly explained.

@@

\begin{figure}
	\includegraphics[width=0.99\linewidth]{z_img_seismicity}
	\caption{Brazilian seismicity from 1767 to 2014. Source: \citep{bsb_2014}.}
	\label{fig_seismicity}
\end{figure}

@@

\begin{figure}
	\begin{center}
	\begin{table}[H]
		%\begin{tabular}{ p{0.5\textwidth} p{0.5\textwidth} }
		\begin{tabular}{ c c }
		\footnotesize(a)
		& \includegraphics[width=0.70\textwidth]{z_img_hmtk_bsb2014_11_decluster}
		\\
		\footnotesize(b)
        & \includegraphics[width=0.60\textwidth]{z_img_hmtk_bsb2014_11_clusters}
		\end{tabular}
	\end{table}
	\caption{Decluster evaluation (a) using distinct algorithms and windows. There is no significant differences between final results. The map (b) of clusters from selected method shows places where earthquake sequences tends to occur more frequently.}
	\label{fig_decluster}
	\end{center}
\end{figure}

>>>>

Brazil occupies a large stable continental region with seismicity rate lower than Central and Eastern North America, Australia, India or Northern Europe \citep{giardini_1999,johnston_1996a}. Spanning most of the central region of the South American plate, Brazil has earthquakes above magnitude 6 $mb$ with a return period of 50 years or so. Historical records are very short with almost no data before the XX century. Historical evidence of magnitude a 7 has been reported in the Amazon \citep{berrocal_1984,veloso_2014}. However, as in other intraplate settings, seismic hazard is not negligible due to the low attenuation in the cold and thick lithosphere (magnitude 5 $mb$ can be felt up to ~500 km away \citep{assumpcao_et_al_2014}) and special seismic hazard studies have been required for critical facilities such as nuclear power plants \citep{almeida_etal_2013}.

Few studies related to seismic hazard, such as \citet{berrocal_1996,santos_2010} for SE and NE Brazil, have been carried out so far and the first Brazilian seismic building code \citep{nbr_15421_2006,santos_lima_2008} was based on the GSHAP \citep{giardini_1999} results for South America \citep{tanner_shepherd_1997, shedlock_tanner_1999}. Brazilian seismicity is concentrated in several seismic zones \citep{berrocal_1984, assumpcao_et_al_2014}. Except for a peak of activity in Northeast Brazil, both the GSHAP model and the \citet{nbr_15421_2006} zoning (even when the Brazilian recommended levels are in line with those of other nations with higher occurrence of earthquakes \citep{santos_2013}) have not taken into proper account some of the other highly active seismic zones, such as the Porto dos Gaúchos, Mato Grosso state, with recurrent activity since the occurrence of the largest recorded Brazilian event with magnitude 6.2 $mb$ in 1955 \citep{barros_2009}, or the Central Amazon region where concentration of events in the last 50 years is consistent with the location of the possible historical earthquake of magnitude 7 \citep{veloso_2014}. For this reason, it is important to review and update the hazard estimates for Brazil, especially since the number of Brazilian earthquake records has been increased in the last three decades. %the increasing number of Brazilian earthquake records in the last three decades. 
%In addition, the results presented here are part of the \acrfull{gem} supported SARA project (South America Risk Assessment).

@@

Data on intraplate seismicity in Brazil is currently the Brazilian Seismic Bulletin maintained mainly by the Seismological Center of the University of São Paulo. This catalog is a joint compilation from different sources \citep{bsb_2014}.

@@

For events up to 1981, the catalog corresponds to \citet{berrocal_1984} which is a comprehensive compilation of historic and instrumental data with participation from Universidade de São Paulo (USP), Universidade de Brasilia (UnB), Universidade Federal do Rio Grande do Norte (UFRN), Observatório Nacional (ON) and Instituto de Pesquisas Tecnológicas do Estado de São Paulo (IPT). From 1982 to 1995 the information comes from the annual bulletins published by Revista Brasileira de Geofísica. Since 1995, the Bulletin is maintained mainly by USP under the same cooperation network in addition to Universidade Estadual Paulista (UNESP). Earlier versions of this catalog were used to compose the well-known \citet{ceresis_1985, ceresis_1995} catalogs.

@@

This catalog only cover crustal events. Andean subduction earthquakes deeper than 50 km in the Brazil-Peru border was discarded as well earthquakes from the Mid-Atlantic ridge. Epicenters into the catalog are shown in figure \ref{fig_seismicity}.

@@

Most magnitudes were computed as $mb$ type or the equivalent $mR$ \citep{assumpcao_1983} regional magnitude. For historical events magnitudes were estimated from macroseismic data (\glsdesc{sym:Af}, \gls{sym:Af} and \glsdesc{sym:I0}, \gls{sym:I0}) \citep{berrocal_1984}.

Since most \glspl{gmpe} are based on $M_W$ moment scale magnitude, in this present study, we converted all $mb$ magnitudes to $M_W$ values following the guidelines on table \ref{tab:guidelines}. REFERENCIA PARA STEPHANE/ASSUMPCAO WORK IN PROGRESS. (Johnston)

Other proxies for intraplate regions like \citet{scordilis_2006} was discarded by its magnitude range definition, over than 6, whilst most magnitudes on \citet{bsb_2014} catalog had values lower than this.

@@

Figures \ref{fig_overview}b and \ref{fig_overview}c show an almost uniform distribution of day of the week, and a slightly trend to record earthquakes during the night hours. This could suggest some influence of daily noise level on the station records.

@@

In the case of Brazilian catalog the methods did not present large differences at the final number of earthquakes (fig. \ref{fig_decluster}a) and the Gardner-Knopoff/Uhrhammer combination was chosen to preserve the maximum number of events.

The map of figure \ref{fig_decluster}b locates all clusters in the catalog. It is an \emph{a-priori} evidence of regions in Brazil where  earthquake sequences tend to occur more frequently.

@@

Brazil just recently installed its permanent seismographic network \citep{pirchiner_2011}. The number and quality of stations increased over time. The population density, previously concentrated along the coast increased also on other regions changing the historical network coverage. %Trace back this history is a huge work with the current available station metadata, mainly from the analog stations.
There are many methodologies to assessment of the temporal and spatial completeness \citep{stepp_1972, mignan_woessner_2012, ogata_1993, wiemer_wyss_2000, cao_gao_2002, stucchi_2004, woessner_2005, mignan_2011, mignan_2012, vorobieva_2013, mignan_2013, nasir_2013, mignan_2014}, or explaining how to compute the real \gls{pde} from the waveform records \citep{schorlemmer_2008}, but the time and efforts to proceed this exhaustive study of the \glsdesc{sym:Mc}, despite its relevance is beyond the scope of this work. The waveforms from earliest years was not in digital form. The histories of most regional (temporary) stations are not easy available. In addition, the low number of earthquakes in the catalog also difficult some high resolution statistical methods. All of this hinders the direct application of most that methodologies. For this reason, the spatial completeness was not considered and just simple temporal \gls{sym:Mc} parameters for the whole country was defined by an easy cumulative rate criteria using historical information about the records evolution and the \citet{stepp_1972} plot (fig. \ref{fig_completeness}a).

Figure \ref{fig_completeness}b shows the magnitude by time distribution in a scatter plot as well the cumulative number of events which is important to perform the \gls{tcef} \citep{nasir_2013}. The linear behavior of cumulative rate on the recent years, gives an idea about the constant record trend. For magnitudes greater than 3, this threshold occurs about 1970. For the magnitudes greater than 4 and 5, these limits could be extrapolated to 1950 and 1940, respective and roughly. Figure \ref{fig_completeness}b also show the completeness tables from the \citet{stepp_1972} test using $\Delta_{mag}=0.5$ and $\Delta_{time}=2$ as magnitude and time bins, and from the literature \citep{assumpcao_et_al_2014}. Tables \ref{tab:completeness_stepp} and \ref{tab:completeness} show each one of these completeness tables.

%------------------------------------------------
%
% smooothing methods
%
%------------------------------------------------
\section{Smoothing methods}
The smoothing methods considered in this article are now briefly explained.

@@

\begin{figure}
	\includegraphics[width=0.99\linewidth]{z_img_seismicity}
	\caption{Brazilian seismicity from 1767 to 2014-01-01. Source: \citep{bsb_2014}.}
	\label{fig_seismicity}
\end{figure}

@@

\begin{figure}
	\begin{center}
	\begin{table}[H]
		%\begin{tabular}{ p{0.5\textwidth} p{0.5\textwidth} }
		\begin{tabular}{ c c }
		\footnotesize(a)
		& \includegraphics[width=0.70\textwidth]{z_img_hmtk_bsb2014_11_decluster}
		\\
		\footnotesize(b)
        & \includegraphics[width=0.60\textwidth]{z_img_hmtk_bsb2014_11_clusters}
		\end{tabular}
	\end{table}
	\caption{Decluster evaluation (a) using distinct algorithms and windows. There is no significant differences between final results. The map (b) of clusters from selected method shows places where earthquake sequences tends to occur.}
	\label{fig_decluster}
	\end{center}
\end{figure}
